Covariate Analysis of ROSMAP Reprocessed counts

### 2020-12-22 code/2020_12_09_ROSMAP_RinEQGt5Cohort_Regress_hg38.Rmd
recommit
rerun

@@

#Filter for Rin Here#
Meta[ Meta$RIN >= 5 | Meta$sampleid %in% c('380_120503','500_120515'), ]

@@

### Filter genes
* Remove genes that have less than 1 cpm counts in at least 50% of samples per Diagnosis
* Remove genes with missing gene length and percentage GC content
```{r filter.genes}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Tissue.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)
genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gene_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene_length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous"))
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=6, fig.width=18, cache=FALSE}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$Tissue.Diagnosis.SEX = paste(COVARIATES$Tissue.Diagnosis,COVARIATES$sex, sep = '.') %>% factor
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('Tissue.Diagnosis.SEX', setdiff(postAdjustCovars, 'SEX')), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
# Re-estimate voom weights
VOOM.WEIGHTS = voom(cnts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)
cntr = c('Tissue.Diagnosis.SEXACC.AD.male-Tissue.Diagnosis.SEXACC.CONTROL.male',
         'Tissue.Diagnosis.SEXACC.AD.female-Tissue.Diagnosis.SEXACC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXACC.AD.male+0.5*Tissue.Diagnosis.SEXACC.AD.female
         -0.5*Tissue.Diagnosis.SEXACC.CONTROL.male-0.5*Tissue.Diagnosis.SEXACC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXACC.AD.female+0.5*Tissue.Diagnosis.SEXACC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXACC.AD.male-0.5*Tissue.Diagnosis.SEXACC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXACC.AD.female-0.5*Tissue.Diagnosis.SEXACC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXACC.AD.male+0.5*Tissue.Diagnosis.SEXACC.CONTROL.male',
         'Tissue.Diagnosis.SEXDLPFC.AD.male-Tissue.Diagnosis.SEXDLPFC.CONTROL.male',
         'Tissue.Diagnosis.SEXDLPFC.AD.female-Tissue.Diagnosis.SEXDLPFC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXDLPFC.AD.male+0.5*Tissue.Diagnosis.SEXDLPFC.AD.female
         -0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.male-0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXDLPFC.AD.female+0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXDLPFC.AD.male-0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXDLPFC.AD.female-0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXDLPFC.AD.male+0.5*Tissue.Diagnosis.SEXDLPFC.CONTROL.male',
         'Tissue.Diagnosis.SEXPCC.AD.male-Tissue.Diagnosis.SEXPCC.CONTROL.male',
         'Tissue.Diagnosis.SEXPCC.AD.female-Tissue.Diagnosis.SEXPCC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXPCC.AD.male+0.5*Tissue.Diagnosis.SEXPCC.AD.female
         -0.5*Tissue.Diagnosis.SEXPCC.CONTROL.male-0.5*Tissue.Diagnosis.SEXPCC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXPCC.AD.female+0.5*Tissue.Diagnosis.SEXPCC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXPCC.AD.male-0.5*Tissue.Diagnosis.SEXPCC.CONTROL.female',
         '0.5*Tissue.Diagnosis.SEXPCC.AD.female-0.5*Tissue.Diagnosis.SEXPCC.CONTROL.female
         -0.5*Tissue.Diagnosis.SEXPCC.AD.male+0.5*Tissue.Diagnosis.SEXPCC.CONTROL.male')
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c('ACC.AD.MALE-ACC.CONTROL.MALE',
                       'ACC.AD.FEMALE-ACC.CONTROL.FEMALE',
                       'ACC.AD.ALL-ACC.CONTROL.ALL',
                       'ACC.ALL.FEMALE-ACC.ALL.MALE',
                       'ACC.ALL.ALL-ACC.ALL.ALL',
                       'DLPFC.AD.MALE-DLPFC.CONTROL.MALE',
                       'DLPFC.AD.FEMALE-DLPFC.CONTROL.FEMALE',
                       'DLPFC.AD.ALL-DLPFC.CONTROL.ALL',
                       'DLPFC.ALL.FEMALE-DLPFC.ALL.MALE',
                       'DLPFC.ALL.ALL-DLPFC.ALL.ALL',
                       'PCC.AD.MALE-PCC.CONTROL.MALE',
                       'PCC.AD.FEMALE-PCC.CONTROL.FEMALE',
                       'PCC.AD.ALL-PCC.CONTROL.ALL',
                       'PCC.ALL.FEMALE-PCC.ALL.MALE',
                       'PCC.ALL.ALL-PCC.ALL.ALL'
                       )
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)
# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)
DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Tissue.Diagnosis','',Comparison),
                Study = 'ROSMAP',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gene_gc_content, gene_length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
DE$Tissue <- do.call( rbind, strsplit( DE$Comparison, '[.]'))[,1]
DE$Comparison <- gsub('DLPFC.', '',gsub('ACC.', '', gsub('PCC.', '',  DE$Comparison )))
tmp = DE %>%
  dplyr::select(ensembl_gene_id, Tissue, Comparison, Direction) %>%
  group_by(Tissue, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)
DE$Tissue <- do.call( rbind, strsplit( DE$Comparison, '[.]'))[,1]
all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
all.fit = c(all.fit, list(Diagnosis.SEX = FIT))

>>>>

#Toss Zero variance genes post residual correct
COUNT <- COUNT[ (row.names(COUNT) %in% c("ENSG00000239467", "ENSG00000005486", "ENSG00000160972", "ENSG00000165886", "ENSG00000002330", "ENSG00000149823", "ENSG00000139722", "ENSG00000169217", "ENSG00000127774", "ENSG00000122490", "ENSG00000130881", "ENSG00000126005", "ENSG00000171222", "ENSG00000070413", "ENSG00000211459")) == F, ]

@@

#Filter for Rin Here#
Meta <- Meta[ Meta$RIN >= 5 | Meta$sampleid %in% c('380_120503','500_120515'), ]

@@

### Filter genes
* Remove genes that have less than 1 cpm counts in at least 50% of samples per Diagnosis
* Remove genes with missing gene length and percentage GC content
```{r filter.genes}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(Tissue.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                     MIN_GENE_CPM=1, 
                                                     MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
    processed.counts$filteredExprMatrix$genes
  }, COUNT)
#filter out non-variable genes post residual removal

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC$ensembl_gene_id[!is.na(GENE.GC$percentage_gene_gc_content)]) %>%
  intersect(GENE.LEN$ensembl_gene_id[!is.na(GENE.LEN$gene_length)]) %>%
  setdiff(c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous")) %>%
  setdiff( c("ENSG00000239467", "ENSG00000005486", "ENSG00000160972", "ENSG00000165886", "ENSG00000002330", "ENSG00000149823", "ENSG00000139722", "ENSG00000169217", "ENSG00000127774", "ENSG00000122490", "ENSG00000130881", "ENSG00000126005", "ENSG00000171222", "ENSG00000070413", "ENSG00000211459") )

#18404
#vs
#18415
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=6, fig.width=18, cache=FALSE}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$Tissue.Diagnosis.SEX = paste(COVARIATES$Tissue.Diagnosis,COVARIATES$sex, sep = '.') %>% factor
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c('Tissue.Diagnosis.SEX', setdiff(postAdjustCovars, 'SEX')), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
# Re-estimate voom weights
VOOM.WEIGHTS = voom(cnts,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS)

colnames(FIT$coefficients) <- gsub( 'Tissue.Diagnosis.SEX','',colnames(FIT$coefficients))
# Fit contrast
contrast = makeContrasts(contrasts=c("(ACC.AD.female+ACC.AD.male)/2-(ACC.CONTROL.female+ACC.CONTROL.male)/2",
                                     "(ACC.AD.female+ACC.CONTROL.female)/2-(ACC.AD.male+ACC.CONTROL.male)/2",
                                     "(ACC.AD.female-ACC.CONTROL.female)/2-(ACC.AD.male-ACC.CONTROL.male)/2",
                                     "ACC.AD.female-ACC.CONTROL.female",
                                     "ACC.AD.male-ACC.CONTROL.male",
                                     "(DLPFC.AD.female+DLPFC.AD.male)/2-(DLPFC.CONTROL.female+DLPFC.CONTROL.male)/2",
                                     "(DLPFC.AD.female+DLPFC.CONTROL.female)/2-(DLPFC.AD.male+DLPFC.CONTROL.male)/2",
                                     "(DLPFC.AD.female-DLPFC.CONTROL.female)/2-(DLPFC.AD.male-DLPFC.CONTROL.male)/2",
                                     "DLPFC.AD.female-DLPFC.CONTROL.female",
                                     "DLPFC.AD.male-DLPFC.CONTROL.male",
                                     "(PCC.AD.female+PCC.AD.male)/2-(PCC.CONTROL.female+PCC.CONTROL.male)/2",
                                     "(PCC.AD.female+PCC.CONTROL.female)/2-(PCC.AD.male+PCC.CONTROL.male)/2",
                                     "(PCC.AD.female-PCC.CONTROL.female)/2-(PCC.AD.male-PCC.CONTROL.male)/2",
                                     "PCC.AD.female-PCC.CONTROL.female",
                                     "PCC.AD.male-PCC.CONTROL.male"),
                         levels = colnames(FIT$coefficients))
colnames(contrast) = c('ACC.AD.MALE-ACC.CONTROL.MALE',
                       'ACC.AD.FEMALE-ACC.CONTROL.FEMALE',
                       'ACC.AD.ALL-ACC.CONTROL.ALL',
                       'ACC.ALL.FEMALE-ACC.ALL.MALE',
                       'ACC.ALL.ALL-ACC.ALL.ALL',
                       'DLPFC.AD.MALE-DLPFC.CONTROL.MALE',
                       'DLPFC.AD.FEMALE-DLPFC.CONTROL.FEMALE',
                       'DLPFC.AD.ALL-DLPFC.CONTROL.ALL',
                       'DLPFC.ALL.FEMALE-DLPFC.ALL.MALE',
                       'DLPFC.ALL.ALL-DLPFC.ALL.ALL',
                       'PCC.AD.MALE-PCC.CONTROL.MALE',
                       'PCC.AD.FEMALE-PCC.CONTROL.FEMALE',
                       'PCC.AD.ALL-PCC.CONTROL.ALL',
                       'PCC.ALL.FEMALE-PCC.ALL.MALE',
                       'PCC.ALL.ALL-PCC.ALL.ALL'
                       )
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)
# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)
DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Tissue.Diagnosis','',Comparison),
                Study = 'ROSMAP',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gene_gc_content, gene_length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
DE$Tissue <- do.call( rbind, strsplit( DE$Comparison, '[.]'))[,1]
DE$Comparison <- gsub('DLPFC.', '',gsub('ACC.', '', gsub('PCC.', '',  DE$Comparison )))
tmp = DE %>%
  dplyr::select(ensembl_gene_id, Tissue, Comparison, Direction) %>%
  group_by(Tissue, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)
DE$Tissue <- do.call( rbind, strsplit( DE$Comparison, '[.]'))[,1]
all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
all.fit = c(all.fit, list(Diagnosis.SEX = FIT))
```
